Health and Economic Impacts of Weather Events
========================================================

```{r setoptions, echo=FALSE}
 opts_chunk$set(message=FALSE,warning=FALSE)
```


Synopsis
------------------------
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.   

The analysis attempts to answer two key questions:  
1. Across the United States, which types of events are most harmful with respect to population health?  
2. Across the United States, which types of events have the greatest economic consequences?

Data used in the analysis is from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

Analysis
--------------------
### Data Processing

```{r dataProc}
setInternet2(use=TRUE)
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "StormData.csv.bz2")

data<-read.csv(bzfile("StormData.csv.bz2"))
unlink("StormData.csv.bz2")
```

Data Profiling and factor creation:
```{r dataFeatures}
names(data)
summary(data)

data$health = data$FATALITIES + data$INJURIES
data$evtype = toupper(data$EVTYPE)
```

### Results

#### Across the United States, which types of events are most harmful with respect to population health?

For the purposes of this analysis, population health related issues refers to injuries and fatalities.  Please see the below graphic to understand the total count of health incidents per event type from 1950 - 2011.  

```{r}
library(sqldf)
library(ggplot2)

#total the total health related issues by event type, select the top 10 
dataHealth<-head(sqldf("SELECT evtype, SUM(health) as `health` FROM data GROUP BY evtype ORDER BY health DESC"),10)

#Plot the data
 ggplot(data=dataHealth, aes(factor(dataHealth$EVTYPE,levels=dataHealth$EVTYPE), dataHealth$health)) + geom_bar(stat="identity")+ylab("Total injuries + fatalities") +xlab("")+ggtitle("Weather Events 1950-2011: Most Harmful to Health")+theme(axis.text.x = element_text(angle=90, vjust=1))

dataHealth
```
Tornados have caused the most injuries and death since 1950 of all weather related events.  
  

#### Across the United States, which types of events have the greatest economic consequences?

For the purpose of this analysis, property and crop damage is used as the indicator of consequence.  We will count the occurrance of damage due to the weather.

Please see the below graphic to understand the total number of economic impacting weather events from 1950 - 2011.  

```{r}
#create one factor to identify all occurrances of weather events with economic impact

data$Amt<- ((data$PROPDMG + data$CROPDMG)>0)

#show the count of events with damage by event type, select top 10
dataProp<-head(sqldf("SELECT evtype, Count(Amt) as `Damage` FROM data WHERE Amt=1 GROUP BY evtype ORDER BY Damage DESC"),10)

#plot the data
ggplot(data=dataProp, aes(factor(dataProp$EVTYPE,levels=dataProp$EVTYPE), dataProp$Damage)) + geom_bar(stat="identity")+ylab("Count storm with economic damage") +xlab("")+ggtitle("Weather Events 1950-2011: Most economic damage Damage")+theme(axis.text.x = element_text(angle=90, vjust=1))

dataProp
```
The weather events with the greatest count of occurrances with economic impact between 1950-2011 is TSTM WIND.


